---
title: "Lab 09 Replication Notebook"
author: "Christopher Prener, Ph.D."
date: '(`r format(Sys.time(), "%B %d, %Y")`)'
output: 
  github_document: default
  html_notebook: default 
always_allow_html: yes
---

## Introduction
This is the replication notebook for Lab 09 from the course SOC 4650/5650: Introduction to GISc.

## Load Dependencies
The following code loads the package dependencies for our analysis:

```{r package-load}
# tidyverse packages
library(dplyr)   # data wrangling
library(ggplot2) # plot data
library(readr)   # work with csv files

# other packages
library(here)    # manage file paths
library(leaflet) # interactive mapping
library(sf)      # spatial data tools
```

## Load Data
We'll use data distributed with the `lecture-10` repository:

```{r load-data}
capitals <- read_csv(here("data", "stateCapitals.csv"))
counties <- st_read(here("data", "US_BOUNDARY_Counties", "US_BOUNDARY_Counties.shp"), 
                    stringsAsFactors = FALSE)
noins <- read_csv(here("data", "US_HEALTH_noIns.csv"))
```

## Projecting State Capitals
The state capitals data contain four variables and a row for each of the 50 U.S. states:

```{r glimpse-capitals}
glimpse(capitals)
```

The `longitude` data is always the `x` variable and the `latitude` data is always the `y` variable. To project based on these two variables, we'll use the `st_as_sf()` function and assign the output to a new object:

```{r project-capitals}
capitals_sf <- st_as_sf(capitals, coords = c(x = "longitude", y = "latitude"), crs = 4326)
```

Note that we used the `WGS 1984` coordinate system initially so that we don't get errors from `leaflet`. We'll use `leaflet` to test our projected point data and make sure that the markers appear over state capitals:

```{r preview-capitals}
leaflet() %>%
  addTiles() %>%
  addMarkers(data = capitals)
```

With our data visually verified, we can go ahead and (1) re-project them from `WGS 1984` to `NAD 1983` and then (2)  export them to our project's `data/` directory.

```{r write-capitals}
capitals_sf %>%
  st_transform(crs = 4269) %>%
  st_write(here("data", "US_CAPITALS", "US_CAPITALS.shp"), delete_dsn = TRUE)
```

Notice how we can pipe these functions - the `sf` package allows us to use `%>%` to pass `sf` object names through functions. Since we are using a function that writes to our hard disk, we do not need to include an outgoing assignment to an object at the end of the pipe.

## Create Counties Shapefile
We loaded data about county boundaries and un-insurance rates earlier in the notebook. It is time to join these data so that we have a combined shapefile. First, we have to check to make sure we have compatible ID variables:

```{r check-for-id}
glimpse(counties)
glimpse(noins)
```

The good news is we have a `GEOID` variable in both tables. The bad news is that they are different forms - the insurance data is formatted as character or string. We can fix this using `mutate()`:

```{r convert-geoid-string}
noins <- mutate(noins, GEOID = as.numeric(GEOID))
```

We'll then use the `left_join()` function from `dplyr` to combine them:

```{r left-join}
noins_sf <- left_join(counties, noins, by = "GEOID")
```

We can check to make sure this worked using `glimpse()`:

```{r gimplse-join}
glimpse(noins_sf)
```

All looks good - we have have both the `noIns` variable and the `geometry` column, so we should be in business. We do have a bit of redundant data here, and could therefore drop by `state` and `county` variables from the data to save on storage space:

```{r drop-vars}
noins_sf <- select(noins_sf, -state, -county)
```

We also need to remove the deliberately coded missing values, which appear as `-1` in the data:

```{r remove-missing}
noins_sf <- filter(noins_sf, noIns >= 0)
```


Finally, we'll confirm that we are using `NAD 1983` (where the `EPSG` value is `4269`):

```{r}
st_crs(noins_sf)
```

Since no `EPSG` code is set, we'll add one. To do this, we'll use `st_set_crs()` instead of `st_transform()`:

```{r set-crs}
noins_sf <- st_set_crs(noins_sf, 4269)
```

We'll check to make sure the addition worked:

```{r check-new-crs}
st_crs(noins_sf)
```

Excellent! Now that they are properly projected, we can quickly preview these data as well using `ggplot2`:

```{r preview-ins-data}
noins_sf %>%
  filter(STATEFP != "02" & STATEFP != "15" & STATEFP != "72") %>%
  ggplot() +
    geom_sf(mapping = aes(fill = noIns), color = NA) +
    scale_fill_distiller(palette = "RdPu", trans = "reverse")
```

Finally, we'll write these data to a shapefile. Before excuting this step, I add a folder to the `data/` directory. You can also do this first step manually if you'd like.

```{r}
dir.create(here("data", "US_HEALTH_noIns"))
st_write(noins_sf, here("data", "US_HEALTH_noIns", "US_HEALTH_noIns.shp"), delete_dsn = TRUE)
```
