---
title: "Lab 09 Replication Notebook"
author: "Christopher Prener, Ph.D."
date: '(`r format(Sys.time(), "%B %d, %Y")`)'
output: 
  github_document: default
  html_notebook: default 
always_allow_html: yes
---

## Introduction
This is the replication notebook for Lab 09 from the course SOC 4650/5650: Introduction to GISc.

## Load Dependencies
The following code loads the package dependencies for our analysis:

```{r package-load}
# tidyverse packages
library(dplyr)   # data
library(readr)   # work with csv files

# other packages
library(here)    # manage file paths
library(leaflet) # interactive mapping
library(sf)      # spatial data tools
```

## Load Data
We'll use data distributed with the `lecture-10` repository:

```{r load-data}
capitals <- read_csv(here("data", "stateCapitals.csv"))
```

## Projecting State Capitals
The state capitals data contain four variables and a row for each of the 50 U.S. states:

```{r}
glimpse(capitals)
```

The `longitude` data is always the `x` variable and the `latitude` data is always the `y` variable. To project based on these two variables, we'll use the `st_as_sf()` function and assign the output to a new object:

```{r}
capitals_sf <- st_as_sf(capitals, coords = c(x = "longitude", y = "latitude"), crs = 4326)
```

Note that we used the `WGS 1984` coordinate system initially so that we don't get errors from `leaflet`. We'll use `leaflet` to test our projected point data and make sure that the markers appear over state capitals:

```{r}
leaflet() %>%
  addTiles() %>%
  addMarkers(data = capitals)
```

With our data visually verified, we can go ahead and (1) re-project them from `WGS 1984` to `NAD 1983` and then (2)  export them to our project's `data/` directory.

```{r}
capitals_sf %>%
  st_transform(crs = 4269) %>%
  st_write(here("data", "US_CAPITALS", "US_CAPITALS.shp"), delete_dsn = TRUE)
```

Notice how we can pipe these functions - the `sf` package allows us to use `%>%` to pass `sf` object names through functions. Since we are using a function that writes to our hard disk, we do not need to include an outgoing assignment to an object at the end of the pipe.

