---
title: "Lecture 10 Notebook"
author: "Christopher Prener, Ph.D."
date: '(`r format(Sys.time(), "%B %d, %Y")`)'
output: 
  github_document: default
  html_notebook: default 
always_allow_html: yes
---

## Introduction
This is the lecture notebook for Lecture-10 from the course SOC 4650/5650: Introduction to GISc.

## Load Dependencies
The following code loads the package dependencies for our analysis:

```{r package-load}
library(leaflet) # interactive mapping
library(sf)    # spatial data tools
```

## Load Data
We'll use two data sets from the `stlData` package to practice projecting data and working with coordinate systems:

```{r load-data}
library(stlData)
stl <- stl_sf_boundary
campus <- stl_tbl_sluPlaces
```

## Working with Coordinate Systems
To begin, we'll explore the projection data for the `stl` object:

```{r preview-city}
stl
```

We can see that the `epsg` value is `4269`. When we look at the `proj4string` value, we see a reference to the `NAD83` datum. If we [look up the given `epsg` value of `4269`](http://spatialreference.org/ref/epsg/4269/), we will see that it is for the `NAD 1983` geographic coordinate system. This gives us a starting place. 

It is also possible to get just the top portion of the output above that tells us about the projection system used by the `sf` object using the `st_crs()` function:

```{r crs-only}
st_crs(stl)
```

If we needed to re-project the data using a different coordinate system, we can use `st_transform()` to do so. For instance, we could convert the `stl` data to WGS 1984 using the epsg code `4326`:

```{r transform-city-wgs84}
stl_wgs <- st_transform(stl, crs = 4326)
```

With the change made, we can again check the object's metadata:

```{r check-city-wgs84}
stl_wgs
```

## Projecting x,y Data
We can also use `sf` to convert tabular data that contains x,y coordinates to a simple features object. We need to have a sense of a few important points. First, we need to know the names of the `x` and `y` coordinate variables:

```{r preview-campus-data}
campus
```

Longitude, or `lng` in this case, is *always* the `x` coordinate. Latitude, or `lat` in this case, is *always* the `y` coordinate.

We also need to know the coordinate system that was used to create the `x` and `y` coordinate columns. This can be tricky to ascertain. The data above are easy to identify as decimal degrees data, but it isn't always quite that simple. For decimal degrees data, we can project them using a geographic coordinate system. We typically use NAD 1983 for this in the U.S., but WGS 1984 works as well. 

### Completing the Projection
We'll use the `epsg` value `4269` for NAD 1983 here to actually project the data using the `st_as_sf()` function:

```{r project-campus-data}
campus_sf <- st_as_sf(campus, coords = c("lng", "lat"), crs = 4269)
```

The `st_as_sf()` function takes the data, our specification for the `crs`, and the given `x` and `y` coordinates and returns a simple features object:

```{r check-campus-projection}
campus_sf
```

We now get both the geometric data and metadata when we print the `campus_sf` object.

### Checking the Project
Given the uncertainty that often exists with projecting x,y data, we should always check our projects (see the lecture slides for some examples of how this can go wrong!). We'll use `leaflet`, the interactive mapping tool, to check our data. The interactivity and easy-to-read basemaps help tremendously in checking our work, and `leaflet` is preferred over `ggplot2` for this reason.

```{r leaflet-map}
leaflet() %>%
  addTiles() %>%
  addMarkers(data = campus_sf)
```

Excellent! Our data project and appear to be on campus. This is a good sign. If they appeared out at sea or in a different part of the country, we would need to check that the columns and the `crs` value were specified correctly.

We did, however, get a warning from `leaflet()` about the layer having an "inconsistent datum". Leaflet wants all of the data mapped to be in WGS 1984. We can easily rectify this using `st_transform()`:

```{r reproject-to-wgs84}
campus_sf_wgs <- st_transform(campus_sf, crs = 4326)
```

With that change made, we can make our interactive map again without any warnings:

```{r leaflet-map-reprojected}
leaflet() %>%
  addTiles() %>%
  addMarkers(data = campus_sf_wgs)
```

